---
title: "Homework 2"
author: "Anika Shareef"
date: "2025-01-27"
output:
  pdf_document: default
  html_document:
    df_print: paged
editor_options:
  markdown:
    wrap: 72
---

EID: as236452

<https://github.com/anikashareef/sds315_hw2.git>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, echo=FALSE}
library(ggplot2)
library(dplyr)
library(tidyverse)
library(knitr)
library(lubridate)

```

## Problem 1: Beauty, or not, in the classroom

#### Part A.

**Create a histogram to display the overall data distribution of course
evaluation scores.**

```{r, message=FALSE, echo=FALSE,fig.width=6, fig.height=4}

#load data set
profs <- read_csv("profs.csv")

#create plot to display overall data distribution of eval scores
profs|>
ggplot() +
  geom_histogram(aes(
    x=eval),
     bins=30,color='black', fill='paleturquoise3')+
    labs(title="Distribution of Course Evaluation Scores",
         x="Scores",
         y="Frequency") +
  theme_minimal()

```

This graph displays the distribution of course evaluation scores among
all the professors in the data set. It is a little skewed left, with an
average score of approximately 4.

#### Part B.

**Use side-by-side boxplots to show the distribution of course
evaluation scores by whether or not the professor is a native English
speaker.**

```{r, message=FALSE, echo=FALSE,fig.width=6, fig.height=4}

#create plot to show distribution eval score based on native english or not
profs|>
  ggplot() +
  geom_boxplot(aes(x=native, y=eval), fill='paleturquoise3')+
    labs(
      title="Course Evaluation Scores: Non-Native vs. Native English Speakers",
      x="Native English Speaker or No",
      y="Score")+
  theme_minimal()

```

In this graph the distribution of course evaluation scores are shown
between native and non-native English speakers is shown. It appears that
native English speaking professors have a higher mean score, but a wider
range to that of non-native English speakers.

#### Part C.

**Use a faceted histogram with two rows to compare the distribution of
course evaluation scores for male and female instructors.**

```{r, message=FALSE, echo=FALSE,fig.width=6, fig.height=4}

#create faceted histogram with two rows to compare distribution of course eval scores for male and female instructors 
profs|>
  ggplot()+
  geom_histogram(aes(x=eval), bins=30, fill='paleturquoise3')+
  facet_wrap(~gender, nrow=2)+
  labs(
    title="Course Evaluation Scores: Female vs. Male",
    x= "Score",
    y= "Frequency")+
  theme_minimal()

```

In these two graphs, the distribution for course evaluation score is
shown between Female and Male professors. When comparing the two, we can
see that the female distribution is generally has a more standard
distribution with male professors having a distribution skewed left,
towards the higher scores.

#### Part D.

**Create a scatterplot to visualize the extent to which there may be an
association between the professor’s physical attractiveness (x) and
their course evaluations (y).**

```{r, message=FALSE, echo=FALSE,fig.width=6, fig.height=4}

profs|>
ggplot()+
  geom_point(aes(x=beauty, y=eval), color='paleturquoise4')+
  labs( 
    title="Professor's Phsyical Attractiveness vs Course Evaluation Score",
    y="Score",
    x="Avg. Rating of Physical Attractiveness")+
  theme_minimal()


```

This scatter-plot visualizes the relationship between course evaluation
score and the average rating of a professor's physical attractiveness.
With a correlation coefficient of 0.19, there is very little to no
correlation between these two variables.

## Problem 2: bike sharing

#### Plot A:

**a line graph showing average hourly bike rentals (total) across all
hours of the day (hr).**

```{r, message=FALSE, echo=FALSE,fig.width=6, fig.height=4}

#load data set 
bikeshare <- read_csv("bikeshare.csv")

#create subset
avg_hr <- bikeshare |>
  group_by(hr) |>
  summarize(avgRentals= mean(total))

#create plot to show hourly bike rentals across hours of the day
avg_hr|>
  ggplot()+
  geom_line(aes(y=avgRentals, x=hr), color='paleturquoise3')+
  labs(
    title="Average Hourly Bike Rentals Across All Hours of the Day",
    y= "Avg # of Bike Rentals",
    x="Hour"
  )+
  theme_minimal()


```

This line graph shows the average hourly bike rentals across all hours
of the day. According to the distribution, the average number of bike
rentals peak at approximately hour 8 and 16, and the lowest number being
hour 4.

#### Plot B:

**a faceted line graph showing average bike rentals by hour of the day,
faceted according to whether it is a working day (workingday).**

```{r, message=FALSE, echo=FALSE,fig.width=6, fig.height=4}

#create new subset that has working day vs non working day within each hour of the day
avg_working_day <- bikeshare |>
  group_by(hr, workingday) |>
  summarize(avgRentals = mean(total))

#create plot   
ggplot(data=avg_working_day)+
  geom_line(aes(x=hr, y=avgRentals), color='paleturquoise')+
         facet_wrap(~workingday, labeller = as_labeller(c("1"= "Working Day", "0"= "Not a Working Day")))+
  labs(
    title= "Avg Hourly Bike Rentals by Working Day or Not",
    x="Hour",
    y="Avg # of Bike Rentals"
  )+
  theme_minimal()
  


```

In this plot, we can see the average number of hourly bike rentals
organized by whether its a working day or not. The graph for the non
working days shows a more steady distribution with only 1 peak at hour
13, versus the graph for the working day that has multiple peaks at
around hour 7 and hour 16/17.

#### 

Plot C:

**a faceted bar plot showing average ridership (y) during the 9 AM hour
by weather situation code (weathersit, x), faceted according to whether
it is a working day or not. (Remember that you can focus on a specific
subset of rows of a data set using filter.)**

```{r, message=FALSE, echo=FALSE,fig.width=6, fig.height=4}

#create subset 
ridership_9 <- bikeshare |>
  filter(hr==8)|>
  group_by(weathersit, workingday) |>
  summarize(avgRentals = mean(total))
  
#create plot
ggplot(data=ridership_9)+
  geom_bar(aes(x=factor(weathersit),y= avgRentals, fill=factor(weathersit)),
           stat="identity",
           color='paleturquoise')+
  facet_wrap(~workingday, labeller= as_labeller(c("1"= "Working Day", "0"= "Not a Working Day")))+
  labs(
    title = "Average Ridership During 9 AM by Weather Situation",
    x = "Weather Situation Code",
    y = "Average Rentals",
    fill = "Weather Code"
  ) +
  theme_minimal()

```

This graph shows the average ridership at 9 am by weather situation and
whether its a working day or not. 1 means clear skies, 2 means cloudy
with some mist, and 3 means light snow/light rain. It appears that among
the not working days, there is much less ridership, peaking at just over
100, and not much variation between the type of weather. In the working
day graph, there is much higher ridership, peaking at just over 500, and
a large gap between weather code 1 and 3.

## Problem 3 - Capital Metro UT Ridership

#### 1.

**One faceted line graph that plots average boardings by hour of the
day, day of week, and month. Youshould facet by day of week. Each facet
should include three lines of average boardings (y) by hour ofthe day
(x), one line for each month and distinguished by color. Give the figure
an informative caption in which you explain what is shown in the figure
and also address the following questions, citing evidence from the
figure. Does the hour of peak boardings change from day to day, or is it
broadly similar across days? Why do you think average boardings on
Mondays in September look lower, compared to other days and months?
Similarly, why do you think average boardings on Weds/Thurs/Fri in
November look lower? (Hint: wrangle first, then plot.)**

```{r, message=FALSE, echo=FALSE,fig.width=6, fig.height=4}

#load data set 
capmetro <- read_csv("capmetro_UT.csv")

#create subset 
avg_boarding <- capmetro |>
  group_by(day_of_week, hour_of_day, month) |>
  summarize(avg_hourly_boarding= mean(boarding), .groups = "drop")|>
  mutate(day_of_week = factor(day_of_week, levels = c("Sun","Mon","Tue","Wed","Thu", "Fri", "Sat")))|>
  mutate(month=factor(month,levels=c("Sep","Oct","Nov")))

#create plot 
ggplot(data=avg_boarding)+
  geom_line(aes(y=avg_hourly_boarding, x=hour_of_day, color=factor(month)))+
  facet_wrap(~day_of_week)+
  labs(
    title="Average Boarding by Hour of the Day, Day of Week, & Month",
    y="Average # of Boardings",
    x="Hour",
    color="Month"
  )+
  theme_minimal()

```

This figure displays the average number of boardings by each hour,
separated by day, and between the months of September, October, and
November. It shows that Sunday and Saturday have the least average
number of boardings across all months, while days like Monday and
Wednesday have the highest. The hour of peak boarding remains similar
across weekdays, landing at around hour 15-17. It is likely that the
average boardings are less in September on Mondays across all days
because September marks the end of summer, when many people return back
to school on the first day of the week, Monday. It's also possible that
the average boardings on Wednesday, Thursday, and Friday in Novembers
becuase the weather gets very cold and people go home on Thanksgiving
break.

#### 2.

**One faceted scatter plot showing boardings (y) vs. temperature (x),
faceted by hour of the day, and with points colored in according to
whether it is a weekday or weekend. Give the figure an informative
caption in which you explain what is shown in the figure and also answer
the following question, citing evidence from the figure. When we hold
hour of day and weekend status constant, does temperature seem to have a
noticeable effect on the number of UT students riding the bus?**

```{r, message=FALSE, echo=FALSE,fig.width=6, fig.height=4}

#create subset 
boarding_temp <- capmetro |>
  mutate(weekend_or_weekday= "Weekday")|>
  mutate(weekend_or_weekday=replace(weekend_or_weekday, day_of_week=="Sat"|day_of_week=="Sun", "Weekend"))|>
  group_by(day_of_week,temperature, hour_of_day)

#create plot P
ggplot(data=boarding_temp)+
  geom_point(aes(y=boarding, x=temperature, color=weekend_or_weekday))+
  facet_wrap(~hour_of_day)+
  labs(
    title="Boardings vs. Temperarture by Hour of the Day",
    y="Boardings",
    x="Temperature",
    color="Weekend or Weekday")+
  theme_minimal()



```

This graph shows the number of boardings vs temperature by the hour of
day, organized by Weekends and Weekdays. Generally, there are much fewer
boardings at Hours 6-9 across all days, compared to hours 14-17, which
have much higher boardings, and higher numbers among weekdays. When we
hold hour of day constant and weekend status constant, it appears that
temperature does not have a significant impact on number of boardings.

## Problem 4: Wrangling the Billboard Top 100

#### Part A:

**Make a table of the top 10 most popular songs since 1958, as measured
by the total number of weeks that a song spent on the Billboard Top 100.
Note that these data end in week 22 of 2021, so the most popular songs
of 2021 onwards will not have up-to-the-minute data; please send our
apologies to The Weeknd. Your table should have 10 rows and 3 columns:
performer, song, and count, where count represents the number of weeks
that song appeared in the Billboard Top 100. Make sure the entries are
sorted in descending order of the count variable, so that the more
popular songs appear at the top of the table. Give your table a short
caption describing what is shown in the table. (Note: you’ll want to use
both performer and song in any group_by operations, to account for the
fact that multiple unique songs can share the same title.)**

```{r, message=FALSE, echo=FALSE,fig.width=6, fig.height=4}


#load data set
billboard <- read_csv("billboard.csv")

#create table 
top_10 <- billboard |>
  group_by(performer,song)|>
  summarize(count=n(),.groups="drop") |>
  arrange(desc(count)) |>
  slice_head(n=10) |>
  select(performer, song, count)

kable(top_10, col.names=c("performer", "song", "count"), caption="Top 10 Most Popular Songs Since 1958")


```

This table shows the top 10 most popular songs since 1958. The top
performer and song is Radioactive by Imagine Dragons at 87 weeks, and
the lowest among the ten is Before He Cheats by Carrie Underwood at 64
weeks.

#### Part B:

**Is the “musical diversity” of the Billboard Top 100 changing over
time? Let’s find out. We’ll measure the musical diversity of given year
as the number of unique songs that appeared in the Billboard Top 100
that year. Make a line graph that plots this measure of musical
diversity over the years. The x axis should show the year, while the y
axis should show the number of unique songs appearing at any position on
the Billboard Top 100 chart in any week that year. For this part, please
filter the data set so that it excludes the years 1958 and 2021, since
we do not have complete data on either of those years. Give the figure
an informative caption in which you explain what is shown in the figure
and comment on any interesting trends you see. There are number of ways
to accomplish the data wrangling here. We offer you two hints on two
possibilities:**

```{r, message=FALSE, echo=FALSE,fig.width=6, fig.height=4}

#create subset 
filtered_years <- billboard |>
  filter(year!=1958 & year!=2021)|>
  group_by(year)|>
  summarize(unique=n_distinct(song),.groups="drop")

#create plot 
ggplot(data=filtered_years) +
  geom_line(aes(x=year,y=unique))+
  labs(
    title="Measure of Musical Diversity over the Years",
    x="Year",
    y="# of Unique Songs"
  )+
  theme_minimal()



```

This graph shows the measure of musical diversity over the years. It
shows that the number of unique songs peaked just before 1970, at 800
songs, and was at its lowest at approx. 2002 at just below 400 songs. It
is fairly jagged, but declined after 1970, but spiked in the early
2000's and reached just below 800 songs in 2020.

#### 

Part C:

**Let’s define a “ten-week hit” as a single song that appeared on the
Billboard Top 100 for at least ten weeks. There are 19 artists in U.S.
musical history since 1958 who have had at least 30 songs that were
“ten-week hits.” Make a bar plot for these 19 artists, showing how many
ten-week hits each one had in their musical career. Give the plot an
informative caption in which you explain what is shown.**

```{r, message=FALSE, echo=FALSE,fig.width=6, fig.height=4}

#create subset and wrangle
ten_week<- billboard |>
  group_by(performer,song )|>
  summarize(weeks_on_chart= n(),.groups="drop")|>
  filter(weeks_on_chart>=10)|>
  count(performer, name="ten_week_hits")|>
  filter(ten_week_hits>=30)|>
  arrange(desc(ten_week_hits))

#create plot 
ggplot(data=ten_week)+
  geom_bar(aes(x=performer, y=ten_week_hits), stat="identity")+
  labs(
    title = "Performers with 30+ Songs on the Billboard for 10+ Weeks",
    x = "Performer",
    y = "Number of Songs") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))




```

This bar graph shows the performers that have 30+ songs on the Billboard
for 10+ weeks. Most of the artists appear to be fairly similar in the
number of songs, with the top performer being Elton John, and the other
top artists being Keith Urban, Kenny Chesney, and Madonna.

## 
